---
title: "Analysis of HuggingFace Models Attention"
author: "Sean Trott and Pam Rivière"
date: "November 20, 2024"
output:
  html_document:
    keep_md: yes
    toc: yes
    toc_float: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(dpi = 300, fig.format = "pdf")
```


```{r include=FALSE}
library(tidyverse)
library(lmtest)
library(forcats)
library(broom)
library(lme4)
library(viridis)
library(ggridges)
library(lmerTest)
library(ggrepel)
library(ggcorrplot)
library(Rtsne)

all_colors <- viridis::viridis(10, option = "mako")
my_colors <- all_colors[c(3, 5, 7)]  # Selecting specific colors from the palette
```

# Load Pythia data


```{r}
# setwd("/Users/seantrott/Dropbox/UCSD/Research/Ambiguity/SSD/entangled_meanings/src/analysis/")
setwd("/Users/pamelariviere/Desktop/backburner_projects/projects_sean/entangled_meanings/src/analysis/")
directory_path <- "../../data/processed/rawc/pythia/attention_rs/"
csv_files <- list.files(path = directory_path, pattern = "*.csv", full.names = TRUE)
csv_list <- csv_files %>%
  map(~ read_csv(.))
df_pythia_models <- bind_rows(csv_list) %>%
  mutate(Layer = Layer + 1)

table(df_pythia_models$mpath)
nrow(df_pythia_models)

```

# Entropy over time

## Mean entropy by seed and overall

```{r}
summary_df <- df_pythia_models %>%
  group_by(step, seed_name) %>%
  summarise(
    mean_entropy = mean(Entropy)
  )


summary_avg = summary_df %>%
  group_by(step) %>%
  summarise(
    mean_entropy = mean(mean_entropy)
  )

ggplot(summary_df, aes(x = step, 
                       y = mean_entropy, 
                       color = seed_name)) +
  geom_line(size = .5, alpha=0.6, linewidth=1) +  # Lineplot for mean entropy
  geom_line(data = summary_avg, aes(x = step, y = mean_entropy), 
             color = "black", size = 1.3) + # Smoothed average
  labs(
    title = "",
    x = "Training Step (Log10)",
    y = "Mean Entropy"
  ) +
  theme_minimal() +
  scale_x_log10() +
  scale_y_continuous(limits = c(1.72, 1.85)) +
  theme(text = element_text(size = 15),
        legend.position = "none") +
  geom_vline(xintercept = 1000, 
             linetype = "dotted", 
             size = 1.2) +
  scale_color_manual(values = viridisLite::viridis(9,option = "magma", begin = 0.9, end = 0.15))


```

## Mean entropy by seed, layer, and overall

```{r}
summary_df <- df_pythia_models %>%
  group_by(step, Layer, seed_name) %>%
  summarise(
    mean_entropy = mean(Entropy)
  )


summary_avg = summary_df %>%
  group_by(step, Layer) %>%
  summarise(
    mean_entropy = mean(mean_entropy)
  )

ggplot(summary_df, aes(x = step, 
                       y = mean_entropy, 
                       color = seed_name)) +
  geom_line(size = .5, linewidth = 1, alpha=0.6) +  # Lineplot for mean entropy
  geom_line(data = summary_avg, aes(x = step, y = mean_entropy), 
             color = "black", size = 1.3) + # Smoothed average
  labs(
    title = "",
    x = "Step",
    y = "Mean Entropy"
  ) +
  theme_minimal() +
  scale_x_log10() +
  scale_y_continuous(limits = c(1.72, 1.85)) +
  theme(text = element_text(size = 15),
        legend.position = "none") +
  geom_vline(xintercept = 1000, 
             linetype = "dotted", 
             size = 1.2) +
  scale_color_manual(values = viridisLite::viridis(9,option = "magma", begin = 0.9, end=0.15)) +
  facet_wrap(~Layer)


```


## Mean entropy by layer and seed

```{r}
df_final = df_pythia_models %>%
  filter(step == 143000) %>%
  group_by(seed, seed_name, Layer, Head) %>%
  summarise(mean_entropy = mean(Entropy, na.rm = TRUE),
            sd_entropy = sd(Entropy, na.rm = TRUE),
            count = n()) %>%
  group_by(seed, seed_name) %>%
  mutate(mean_entropy_scaled = scale(mean_entropy)) 

mod_full = lmer(data = df_final, mean_entropy_scaled ~ Layer + 
                  (1 | seed_name) + (1 | Head))

summary(mod_full)


df_final %>%
  ggplot(aes(x = Layer,
             y = Head,
             fill = mean_entropy_scaled)) +
  geom_tile() +
  labs(x = "Layer",
       y = "Head",
       fill = "Mean Entropy (Z-scored)") +
  scale_fill_gradient2(low = "blue",
                       mid = "white",
                       high = "red",
                       # midpoint = 0, 
                       name = "Mean Entropy (Z-scored)") +
  theme_minimal() +
  theme(text = element_text(size = 15),
        strip.text.y = element_text(angle = 0), # Keep facet labels horizontal
        axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "bottom") +
  facet_wrap(~seed_name)

summary_avg = df_final %>%
  group_by(Layer) %>%
  summarise(
    mean_entropy = mean(mean_entropy)
  )


df_final %>%
  group_by(Layer, seed_name) %>%
  summarise(mean_entropy = mean(mean_entropy)) %>%
  ggplot(aes(x = Layer,
             y = mean_entropy, 
             color = seed_name)) +
  geom_line(size = .5, linewidth = 1,alpha=0.5) +  # Lineplot for mean entropy
  geom_line(data = summary_avg, aes(x = Layer, y = mean_entropy), 
             color = "black", size = 3) + # Smoothed average
  labs(
    title = "",
    x = "Layer",
    y = "Mean Entropy"
  ) +
  theme_minimal() +
  theme(text = element_text(size = 15),
        legend.position = "none") +
  scale_color_manual(values = viridisLite::viridis(9,option = "magma", begin=0.9,end=0.15))

```


# Attention


## Attention at final step

```{r}
df_pythia_models %>%
  filter(step == 143000) %>%
  group_by(seed_name, Layer, Head) %>%
  summarise(mean_attention = mean(Attention)) %>%
  ggplot(aes(x = Layer,
             y = Head,
             fill = mean_attention)) +
  geom_tile() +
  labs(x = "Layer",
       y = "Head",
       fill = "Attention to Disambiguating Word") +
  scale_fill_gradient2(low = "blue",
                       mid = "white",
                       high = "red",
                       midpoint = 0, 
                       name = "Attention to Disambiguating Word") +
  theme_minimal() +
  theme(text = element_text(size = 15),
        strip.text.y = element_text(angle = 0), # Keep facet labels horizontal
        axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "bottom") +
  facet_wrap(~seed_name)


df_pythia_models %>%
  filter(step == 143000) %>%
  group_by(seed_name, Layer, Head) %>%
  summarise(mean_attention = mean(Attention)) %>%
  group_by(seed_name, Layer) %>%
  summarise(max_mean = max(mean_attention)) %>%
  ggplot(aes(x = Layer,
             y = seed_name,
             fill = max_mean)) +
  geom_tile() +
  labs(x = "Layer",
       y = "Seed",
       fill = "Max Attn. to Disambiguating Word") +
  scale_fill_gradient2(low = "blue",
                       mid = "white",
                       high = "red",
                       midpoint = 0, 
                       name = "Max Attn. to Disambiguating Word") +
  theme_minimal() +
  theme(text = element_text(size = 15),
        strip.text.y = element_text(angle = 0), # Keep facet labels horizontal
        axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "bottom") 


```

### Clustering by final-step attention

In an exploratory analysis, do these apparent attention patterns naturally separate into two distinct clusters?

```{r}
### Pivot wider
final_step_wide = df_pythia_models %>%
  filter(step == 143000) %>% 
  select(-Entropy, -n_params) %>%
  pivot_wider(names_from = c(Head, Layer),
              values_from = Attention) 

# Get the numeric columns
numeric_columns <- grep("^\\d+_\\d+$", colnames(final_step_wide), value = TRUE)
clustering_data <- final_step_wide %>%
  select(all_of(numeric_columns))

# Scale data for k-means
scaled_data <- scale(clustering_data)

# Run k-means with two clusters
set.seed(123) 
k <- 2
kmeans_result <- kmeans(scaled_data, centers = k, nstart = 25)

## Analyze correspondence to seed
final_step_wide$cluster_kmeans <- kmeans_result$cluster
table(final_step_wide$cluster_kmeans, final_step_wide$seed_name)
round(table(final_step_wide$cluster_kmeans, final_step_wide$seed_name) / 336, 2)


##### Can also visualize using PCA
# Perform PCA
pca_result <- prcomp(scaled_data, center = TRUE, scale. = TRUE)

# Extract first two principal components to plot
final_step_wide$PC1 = pca_result$x[,1]
final_step_wide$PC2 = pca_result$x[,2]

# Plot the first two principal components
final_step_wide %>%
  ggplot(aes(x = PC1, y = PC2, color = seed_name, 
             shape = factor(cluster_kmeans))) +
  geom_point(alpha = .7, size = 2) +
  theme_minimal() +
  geom_vline(xintercept = 0, linetype = "dashed", size = 1.5) +
  labs(x = "PC1", 
       y = "PC2") +
  theme(text = element_text(size = 15),
        legend.position = "none") +
  scale_color_manual(values = viridisLite::viridis(9,option = "magma", begin=0.9,end=0.15))



#### Now do with TSNE
tsne_result <- Rtsne(scaled_data, perplexity = 30, 
                     check_duplicates = FALSE, verbose = TRUE, max_iter = 500)

# Extract the two components and store them in a dataframe
final_step_wide$ts1 = tsne_result$Y[,1]
final_step_wide$ts2 = tsne_result$Y[,2]

# Plot t-SNE results
final_step_wide %>%
  ggplot(aes(x = ts1, y = ts2, color = factor(seed), 
             shape = factor(cluster_kmeans))) +
  geom_point(alpha = .7, size = 2) +
  theme_minimal() +
  geom_vline(xintercept = 0, linetype = "dashed", size = 1.5) +
  labs(x = "Dim 1", 
       y = "Dim 2") +
  theme(text = element_text(size = 15),
        legend.position = "none") +
  scale_color_manual(values = viridisLite::viridis(9,option = "magma", begin=0.9,end=0.15))


```



## Attention over time by head

```{r}
df_attention_over_time = df_pythia_models %>% 
  filter(mpath %in% c("EleutherAI/pythia-14m")) %>%
  group_by(mpath, revision, Layer, Head, step, n_params, seed_name) %>%
  summarise(mean_attention = mean(Attention)) %>%
  mutate(step_modded = step + 1) %>%
  group_by(mpath, step, Head) %>%
  mutate(
    attention_diff = mean_attention - lag(mean_attention),
    Head = Head + 1
  )

df_attention_avg = df_attention_over_time %>% 
  group_by(step_modded, revision, step, Layer, Head) %>%
  summarise(mean_attention = mean(mean_attention))


df_attention_over_time %>%
  filter(Layer == 3) %>%
  ggplot(aes(x = step_modded,
             y = mean_attention,
             color = factor(Head))) +
  geom_point(size = 2, alpha = .7) +
  geom_line(size = 1, alpha = .7) + #, linetype = "dotted") +
  # geom_line(data = df_attention_avg, aes(x = step_modded, y = mean_attention), 
  #           color = "black", size = 1.3) + # Smoothed average
  theme_minimal() +
  labs(x = "Training Step (Log10)",
       y = "Attention to Disambiguating Word",
       color = "Attention Head",
       title = "Layer 3 by Random Seed") +
  scale_x_log10() +
    geom_vline(xintercept = 1000, 
             linetype = "dotted", 
             size = 1.2) +
  theme(text = element_text(size = 15),
        legend.position = "bottom") +
  scale_color_manual(values = viridisLite::viridis(4,option = "mako", begin=0.8,end=0.15)) +
  facet_wrap(~seed_name)


df_attention_over_time %>%
  filter(Layer == 4) %>%
  ggplot(aes(x = step_modded,
             y = mean_attention,
             color = factor(Head))) +
  geom_point(size = 2, alpha = .7) +
  geom_line(size = 1, alpha = .7) + #, linetype = "dotted") +
  # geom_line(data = df_attention_avg, aes(x = step_modded, y = mean_attention), 
  #           color = "black", size = 1.3) + # Smoothed average
  theme_minimal() +
  labs(x = "Training Step (Log10)",
       y = "Attention to Disambiguating Word",
       color = "Attention Head",
       title = "Layer 4 Attention By Random Seed") +
  scale_x_log10() +
    geom_vline(xintercept = 1000, 
             linetype = "dotted", 
             size = 1.2) +
  theme(text = element_text(size = 15),
        legend.position = "bottom") +
  scale_color_manual(values = viridisLite::viridis(4,option = "mako", begin=0.8,end=0.15)) +
  facet_wrap(~seed_name)





```


## Aggregating across layers/heads

```{r}
### Track max disambiguating heads at each time point for each seed, across layer/head
df_by_head_max_attention = df_pythia_models %>%
  mutate(step_modded = step + 1) %>%
  group_by(step_modded, revision, Layer, Head, seed, seed_name) %>%
  summarise(mean_attention = mean(Attention)) %>%
  group_by(step_modded, seed, seed_name) %>%
  slice_max(mean_attention) %>%
  mutate(Head = Head + 1)

summary_avg = df_by_head_max_attention %>%
  group_by(step_modded) %>%
  summarise(
    mean_across_seeds = mean(mean_attention)
  )

df_by_head_max_attention %>%
  ggplot(aes(x = step_modded,
             y = mean_attention,
             color = factor(seed_name))) +
  geom_line(size = .6, linetype = "dotted") +  # Lineplot for mean entropy
  geom_line(data = summary_avg, aes(x = step_modded, y = mean_across_seeds), 
             color = "black", size = 1.5) + # Smoothed average 
  theme_minimal() +
  labs(x = "Training Step (Log10)",
       y = "Max Attn. to Disambiguating Word",
       color = "") +
  scale_x_log10() +
  geom_vline(xintercept = 1000, 
           linetype = "dotted", 
           size = 1.2) +
  theme(text = element_text(size = 15),
        legend.position = "bottom") +
  scale_color_viridis(option = "mako", discrete=TRUE) 


### Track max disambiguation heads at each time point for each seed, across heads
df_by_head_max_attention = df_pythia_models %>%
  mutate(step_modded = step + 1) %>%
  group_by(step_modded, revision, Layer, Head, seed, seed_name) %>%
  summarise(mean_attention = mean(Attention)) %>%
  group_by(step_modded, seed, seed_name, Layer) %>%
  slice_max(mean_attention) %>%
  mutate(Head = Head + 1)

summary_avg = df_by_head_max_attention %>%
  group_by(step_modded, Layer) %>%
  summarise(
    mean_across_seeds = mean(mean_attention)
  )

df_by_head_max_attention %>%
  ggplot(aes(x = step_modded,
             y = mean_attention,
             color = factor(seed_name))) +
  geom_line(size = .6, alpha=0.6, linewidth=2) +  # Lineplot for mean entropy
  geom_line(data = summary_avg, aes(x = step_modded, y = mean_across_seeds), 
             color = "black", size = 3) + # Smoothed average 
  theme_minimal() +
  labs(x = "Training Step (Log10)",
       y = "Max Attn. to Disambiguating Word",
       color = "") +
  scale_x_log10() +
  geom_vline(xintercept = 1000, 
           linetype = "dashed", 
           size = 1.2) +
  theme(text = element_text(size = 15),
        legend.position = "bottom") +
  scale_color_manual(values = viridisLite::viridis(9,option = "magma", begin=0.9,end=0.15))  +
  facet_wrap(~Layer)



df_by_head_max_attention %>%
  filter(Layer == 3) %>%
  ggplot(aes(x = step_modded,
             y = mean_attention,
             color = factor(seed_name))) +
  geom_line(size = .6, alpha=0.6, linewidth=2) +  # Lineplot for mean entropy
  theme_minimal() +
  labs(x = "Training Step (Log10)",
       y = "Max Attn. to Disambiguating Word",
       title = "Layer 3",
       color = "") +
  scale_x_log10() +
  geom_vline(xintercept = 1000, 
           linetype = "dashed", 
           size = 1.2) +
  theme(text = element_text(size = 15),
        legend.position = "bottom") +
  scale_color_manual(values = viridisLite::viridis(9,option = "magma", begin=0.9,end=0.15))  


df_by_head_max_attention %>%
  filter(Layer == 4) %>%
  ggplot(aes(x = step_modded,
             y = mean_attention,
             color = factor(seed_name))) +
  geom_line(size = .6, alpha=0.6, linewidth=2) +  # Lineplot for mean entropy
  theme_minimal() +
  labs(x = "Training Step (Log10)",
       y = "Max Attn. to Disambiguating Word",
       title = "Layer 4",
       color = "") +
  scale_x_log10() +
  geom_vline(xintercept = 1000, 
           linetype = "dashed", 
           size = 1.2) +
  theme(text = element_text(size = 15),
        legend.position = "bottom") +
  scale_color_manual(values = viridisLite::viridis(9,option = "magma", begin=0.9,end=0.15))  


```



