---
title: "Analysis of HuggingFace Models Attention to Previous Tokens"
author: "Sean Trott and Pam Rivière"
date: "November 20, 2024"
output:
  # html_document:
  #  keep_md: yes
  #  toc: yes
  #  toc_float: yes
  pdf_document: 
      fig_caption: yes
      keep_md: yes
      keep_tex: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(dpi = 300, fig.format = "pdf")
```


```{r include=FALSE}
library(tidyverse)
library(lmtest)
library(forcats)
library(broom)
library(lme4)
library(viridis)
library(ggridges)
library(lmerTest)
library(ggrepel)
library(ggcorrplot)

all_colors <- viridis::viridis(10, option = "mako")
my_colors <- all_colors[c(3, 5, 7)]  # Selecting specific colors from the palette
```

# Load Pythia data


Here, we analyze data looking at the average attention each head gives from each token to the previous token.

```{r}
# setwd("/Users/seantrott/Dropbox/UCSD/Research/Ambiguity/SSD/entangled_meanings/src/analysis/")
directory_path <- "../../data/processed/rawc/pythia/attention_1back/"
csv_files <- list.files(path = directory_path, pattern = "*.csv", full.names = TRUE)
csv_list <- csv_files %>%
  map(~ read_csv(.))
df_pythia_models <- bind_rows(csv_list) %>%
  mutate(Layer = Layer + 1,
         Head = Head + 1)

table(df_pythia_models$mpath)

```



# Attention at final step

```{r}
df_final_step = df_pythia_models %>%
  filter(step == 143000)
nrow(df_final_step)

df_attn_by_head = df_final_step %>%
  group_by(mpath, Layer, Head, n_params) %>%
  summarise(mean_attention = mean(Attention)) %>%
  group_by(mpath) %>%
  mutate(max_layer = max(Layer),
         prop_layer = Layer / max_layer) %>%
  mutate(binned_prop_layer = ntile(prop_layer, 6)) %>%
  mutate(prop_binned = binned_prop_layer / 6) 


df_attn_by_head %>%
  ggplot(aes(x = Layer,
             y = Head,
             fill = mean_attention)) +
  geom_tile() +
  labs(x = "Layer",
       y = "Head",
       fill = "Attention to Previous Token") +
  scale_fill_gradient2(low = "blue",
                       mid = "white",
                       high = "red",
                       midpoint = 0, 
                       name = "Attention to Previous Token") +
  theme_minimal() +
  theme(text = element_text(size = 15),
        strip.text.y = element_text(angle = 0), # Keep facet labels horizontal
        axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "bottom") +
  facet_wrap(~reorder(mpath, n_params))


df_attn_by_head %>%
  group_by(mpath, prop_binned, n_params) %>%
  summarise(mean_attention = mean(mean_attention)) %>%
  ggplot(aes(x = prop_binned,
             y = reorder(mpath, n_params),
             fill = mean_attention)) +
  geom_tile() +
  labs(x = "Layer Depth Ratio",
       y = "",
       fill = "Attention to Previous Token") +
  scale_fill_gradient2(low = "blue",
                       mid = "white",
                       high = "red",
                       midpoint = 0, 
                       name = "Attention to Previous Token") +
  theme_minimal() +
  theme(text = element_text(size = 15),
        strip.text.y = element_text(angle = 0), # Keep facet labels horizontal
        axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "bottom") 



df_attn_by_head %>%
  filter(mpath == "EleutherAI/pythia-14m") %>%
  ggplot(aes(x = Layer,
             y = Head,
             fill = mean_attention)) +
  geom_tile() +
  labs(x = "Layer",
       y = "Head",
       fill = "Attention to Previous Token") +
  scale_fill_gradient2(low = "blue",
                       mid = "white",
                       high = "red",
                       midpoint = 0, 
                       name = "Attention to Previous Token") +
  theme_minimal() +
  theme(text = element_text(size = 15),
        strip.text.y = element_text(angle = 0), # Keep facet labels horizontal
        axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "bottom")
```



# Attention over time


## Attention by head

```{r}
df_attention_over_time = df_pythia_models %>% 
  filter(mpath %in% c("EleutherAI/pythia-14m")) %>%
  # filter(Layer == 3) %>%
  group_by(mpath, revision, Layer, Head, step, n_params) %>%
  summarise(mean_attention = mean(Attention)) %>%
  mutate(step_modded = step + 1) %>%
  group_by(mpath, step, Head)

df_attention_over_time %>%
  # filter(Layer == 3) %>%
  ggplot(aes(x = step_modded,
             y = mean_attention,
             color = factor(Head))) +
  geom_point(size = 4, alpha = .7) +
  geom_line(size = 2, alpha = .7) +
  theme_minimal() +
  labs(x = "Training Step (Log10)",
       y = "Attention to Previous Token",
       color = "Attention Head") +
  scale_x_log10() +
    geom_vline(xintercept = 1000, 
             linetype = "dotted", 
             size = 1.2) +
  theme(text = element_text(size = 15),
        legend.position = "bottom") +
  scale_color_viridis(option = "mako", discrete=TRUE) +
  facet_wrap(~Layer)

df_attention_over_time %>%
  filter(Layer == 3) %>%
  # filter(Head %in% c(1, 2)) %>%
  ggplot(aes(x = step_modded,
             y = mean_attention,
             color = factor(Head))) +
  geom_point(size = 4, alpha = .7) +
  geom_line(size = 2, alpha = .7) +
  theme_minimal() +
  labs(x = "Training Step (Log10)",
       y = "Attention to Previous Token",
       color = "Attention Head (Layer 3)") +
  scale_x_log10() +
    geom_vline(xintercept = 1000, 
             linetype = "dotted", 
             size = 1.2) +
  theme(text = element_text(size = 15),
        legend.position = "bottom") +
  scale_color_viridis(option = "mako", discrete=TRUE) 

```





### Attention at sentence-level

```{r}
df_pythia_models %>%
  filter(mpath == "EleutherAI/pythia-14m") %>%
  filter(Layer == 3) %>%
  filter(step %in% c(512, 1000, 2000, 143000)) %>%
  ggplot(aes(x = Attention,
             y = factor(Head),
             fill = factor(mpath))) +
  geom_density_ridges2(aes(height = ..density..), 
                       color=gray(0.25), 
                       alpha = .7, 
                       scale=.85, 
                       # size=1, 
                       size = 0,
                       stat="density") +
  labs(x = "Attention to Previous Token",
       y = "Attention Head",
       fill = "") +
  theme_minimal() +
  scale_fill_viridis(option = "mako", discrete=TRUE) +
  theme(text = element_text(size = 15),
        legend.position="none") +
  facet_wrap(~reorder(revision, step))
```



# Comparing to disambiguating word

## Load initial data

```{r}
# setwd("/Users/seantrott/Dropbox/UCSD/Research/Ambiguity/SSD/entangled_meanings/src/analysis/")
directory_path <- "../../data/processed/rawc/pythia/attention/"
csv_files <- list.files(path = directory_path, pattern = "*.csv", full.names = TRUE)
csv_list <- csv_files %>%
  map(~ read_csv(.))
df_pythia_models_original <- bind_rows(csv_list) %>%
  mutate(Layer = Layer + 1,
         Head = Head + 1) %>%
  mutate(Attention_to_disambiguating = Attention) %>%
  select(-Attention)

table(df_pythia_models_original$mpath)



```


## Merging

```{r}
nrow(df_pythia_models_original)
nrow(df_pythia_models)

df_pythia_all = df_pythia_models %>%
  inner_join(df_pythia_models_original) %>%
  mutate(attention_diff = Attention_to_disambiguating - Attention) %>%
  mutate(attention_ratio = Attention_to_disambiguating / Attention)
nrow(df_pythia_all)
```




## Final step


```{r}

### Calculate
df_attn_by_head_final = df_pythia_all %>%
  filter(step == 143000) %>%
  group_by(mpath, Layer, Head, n_params) %>%
  summarise(attention_diff = mean(attention_diff),
            attention_ratio = mean(attention_ratio)) 



### Difference analysis
df_attn_by_head_final %>%
  ggplot(aes(x = Layer,
             y = Head,
             fill = attention_diff)) +
  geom_tile() +
  labs(x = "Layer",
       y = "Head",
       fill = "Attention Difference") +
  scale_fill_gradient2(low = "blue",
                       mid = "white",
                       high = "red",
                       midpoint = 0, 
                       name = "Attention Difference") +
  theme_minimal() +
  theme(text = element_text(size = 15),
        strip.text.y = element_text(angle = 0), # Keep facet labels horizontal
        axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "bottom") +
  facet_wrap(~reorder(mpath, n_params))


### Difference analysis
df_attn_by_head_final %>%
  ggplot(aes(x = Layer,
             y = Head,
             fill = attention_ratio)) +
  geom_tile() +
  labs(x = "Layer",
       y = "Head",
       fill = "Attention Ratio") +
  scale_fill_gradient2(low = "blue",
                       mid = "white",
                       high = "red",
                       midpoint = 1, 
                       name = "Attention Ratio") +
  theme_minimal() +
  theme(text = element_text(size = 15),
        strip.text.y = element_text(angle = 0), # Keep facet labels horizontal
        axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "bottom") +
  facet_wrap(~reorder(mpath, n_params))

```

## Over time

```{r}
df_attention_over_time = df_pythia_all %>% 
  filter(mpath %in% c("EleutherAI/pythia-14m")) %>%
  group_by(mpath, revision, Layer, Head, step, n_params) %>%
  summarise(mean_attention_1back = mean(Attention),
            mean_attention_disambiguating = mean(Attention_to_disambiguating),
            mean_attention_diff = mean(attention_diff),
            mean_attention_ratio = mean(attention_ratio),
            se_diff = sd(attention_diff) / n()) %>%
  mutate(step_modded = step + 1)

df_attention_over_time %>%
  ggplot(aes(x = step_modded,
             y = mean_attention_diff,
             color = factor(Head))) +
  geom_point(size = 2, alpha = .7) +
  geom_line(size = 2, alpha = .7) +
  theme_minimal() +
  labs(x = "Training Step (Log10)",
       y = "Attention Difference",
       color = "Attention Head") +
  scale_x_log10() +
    geom_vline(xintercept = 1000, 
             linetype = "dotted", 
             size = 1.2) +
    geom_hline(yintercept = 0, 
             linetype = "dashed", 
             size = 1.2,
             color = "black") +
  theme(text = element_text(size = 15),
        legend.position = "bottom") +
  scale_color_viridis(option = "mako", discrete=TRUE) +
  facet_wrap(~Layer)

df_attention_over_time %>%
  filter(Layer == 3) %>%
  ggplot(aes(x = step_modded,
             y = mean_attention_diff,
             color = factor(Head))) +
  geom_point(size = 2, alpha = .7) +
  geom_line(size = 2, alpha = .7) +
  theme_minimal() +
  labs(x = "Training Step (Log10)",
       y = "Attention Difference",
       color = "Attention Head") +
  scale_x_log10() +
    geom_vline(xintercept = 1000, 
             linetype = "dotted", 
             size = 1.2) +
    geom_hline(yintercept = 0, 
             linetype = "dashed", 
             size = 1.2,
             color = "black") +
  theme(text = element_text(size = 15),
        legend.position = "bottom") +
  scale_color_viridis(option = "mako", discrete=TRUE) 

df_attention_over_time %>%
  ggplot(aes(x = step_modded,
             y = mean_attention_ratio,
             color = factor(Head))) +
  geom_point(size = 2, alpha = .7) +
  geom_line(size = 2, alpha = .7) +
  theme_minimal() +
  labs(x = "Training Step (Log10)",
       y = "Attention Ratio",
       color = "Attention Head") +
  scale_x_log10() +
    geom_vline(xintercept = 1000, 
             linetype = "dotted", 
             size = 1.2) +
    geom_hline(yintercept = 1, 
             linetype = "dashed", 
             size = 1.2,
             color = "black") +
  theme(text = element_text(size = 15),
        legend.position = "bottom") +
  scale_color_viridis(option = "mako", discrete=TRUE) +
  facet_wrap(~Layer)

df_attention_over_time %>%
  filter(Layer == 3) %>%
  ggplot(aes(x = step_modded,
             y = mean_attention_ratio,
             color = factor(Head))) +
  geom_point(size = 2, alpha = .7) +
  geom_line(size = 2, alpha = .7) +
  theme_minimal() +
  labs(x = "Training Step (Log10)",
       y = "Attention Ratio",
       color = "Attention Head") +
  scale_x_log10() +
    geom_vline(xintercept = 1000, 
             linetype = "dotted", 
             size = 1.2) +
    geom_hline(yintercept = 1, 
             linetype = "dashed", 
             size = 1.2,
             color = "black") +
  theme(text = element_text(size = 15),
        legend.position = "bottom") +
  scale_color_viridis(option = "mako", discrete=TRUE) 
```


## T-test for subtraction analysis

```{r}
t_test_results <- df_pythia_all %>%
  group_by(mpath, Layer, Head, n_params, step) %>%
  summarise(
    t_test = list(
      t.test(
        Attention_to_disambiguating,
        Attention,
        paired = TRUE,
        alternative = "greater"  # One-tailed test for greater values
      )
    ),
    .groups = "drop"
  ) %>%
  mutate(
    p_value = map_dbl(t_test, ~ .x$p.value),          # Extract p-value
    t_statistic = map_dbl(t_test, ~ .x$statistic)    # Extract t-statistic
  )


# Apply multiple testing correction
t_test_results <- t_test_results %>%
  group_by(mpath) %>%
  mutate(
    p_adjusted = p.adjust(p_value, method = "fdr")  # Adjust p-values (FDR)
  ) %>%
  mutate(p_adj2 = case_when(
    p_adjusted < .05 ~ p_adjusted,
    p_adjusted >= .05 ~ 1
  ))

# Filter significant results
significant_results <- t_test_results %>%
  filter(p_adjusted < 0.05) 


### Significant at final step
significant_results %>%
  filter(step == 143000) %>%
  ggplot(aes(x = Layer, y = Head, fill = -log10(p_adjusted))) +
  geom_tile() +
  scale_fill_gradient(low = "white", high = "red", name = "-log10(FDR p-value)") +
  theme_minimal() +
  labs(
    title = "Significant Heads/Layers",
    x = "Layer",
    y = "Head"
  ) +
  facet_wrap(~reorder(mpath, n_params))



### How many timesteps?
significant_results %>%
  filter(mpath == "EleutherAI/pythia-14m") %>%
  group_by(Layer, Head) %>%
  summarise(count = n(),
            prop = count / 154)

### Geom rug
df_attention_over_time %>%
  filter(mpath == "EleutherAI/pythia-14m") %>%
  inner_join(t_test_results) %>%
  filter(Layer == 3) %>%
  mutate(is_significant = p_adjusted < .05,
         Head = factor(Head)) %>%
  ggplot(aes(x = step_modded,
             y = mean_attention_diff,
             color = Head)) +
  geom_line(size = 1.2, alpha = 0.8) +
  geom_point(alpha = 0.6) +
  # Head-specific rug layer
  geom_rug(data = ~ .x %>% filter(is_significant),
           aes(x = step_modded, color = Head),
           inherit.aes = FALSE,
           sides = "b",
           alpha = 0.8,
           length = unit(0.05, "npc")) +

  scale_x_log10() +
  geom_vline(xintercept = 1000, linetype = "dotted", size = 1.2) +
  geom_hline(yintercept = 0, linetype = "dashed", size = 1.2) +
  scale_color_viridis_d(option = "mako") +
  labs(x = "Training Step (Log10)",
       y = "Attention Difference",
       color = "Head") +
  facet_wrap(~Head, ncol = 2) +  # separate panel per head
  theme_minimal() +
  theme(text = element_text(size = 15),
        legend.position = "none")



### Stacked rugs
df_rugs <- df_attention_over_time %>%
  filter(mpath == "EleutherAI/pythia-14m") %>%
  inner_join(t_test_results) %>%
  filter(Layer == 3, p_adjusted < 0.05) %>%
  mutate(Head = factor(Head),
         rug_ymin = -0.25 - as.numeric(Head) * 0.02,  # more spacing here
         rug_ymax = rug_ymin + 0.01)  
df_attention_over_time %>%
  filter(mpath == "EleutherAI/pythia-14m") %>%
  inner_join(t_test_results) %>%
  filter(Layer == 3) %>%
  mutate(is_significant = p_adjusted < .05,
         Head = factor(Head)) %>%
  ggplot(aes(x = step_modded,
             y = mean_attention_diff,
             color = Head)) +
  geom_line(size = 1.2, alpha = 0.8) +
  geom_point(alpha = 0.5) +

  geom_segment(data = df_rugs,
               aes(x = step_modded, xend = step_modded,
                   y = rug_ymin, yend = rug_ymax, color = Head),
               inherit.aes = FALSE,
               alpha = 0.8) +

  scale_x_log10() +
  geom_vline(xintercept = 1000, linetype = "dotted", size = 1.2) +
  geom_hline(yintercept = 0, linetype = "dashed", size = 1.2) +
  scale_color_viridis_d(option = "mako") +
  labs(x = "Training Step (Log10)",
       y = "Attention Difference",
       color = "Head") +
  theme_minimal() +
  theme(text = element_text(size = 15),
        legend.position = "bottom")
```


## Error analysis

Sentences for which the target head attends on average to previous words more than specifically to the disambiguating word.


```{r}
just_h2 = df_pythia_all %>%
  filter(mpath == "EleutherAI/pythia-14m") %>%
  filter(step == 143000) %>%
  filter(Head == 2) %>%
  filter(Layer == 3)

less_than_zero = just_h2 %>%
  filter(attention_diff < 0) 
```


Anything systematic? (Not related to CS norms for sentences.)

```{r}
df_cs = read_csv("../../../cs_norms/data/processed/contextualized_sensorimotor_norms.csv")

df_merged_cs = df_cs %>%
  inner_join(just_h2, by = c("word", "sentence"))


m = lm(data = df_merged_cs,
       attention_diff ~ Vision.M + Touch.M + Taste.M + Olfaction.M + 
         Hearing.M + Interoception.M +
         Head.M + Mouth_throat.M + Hand_arm.M + Torso.M + Foot_leg.M)

cols = df_merged_cs %>%
  select(attention_diff, Vision.M, Touch.M, Taste.M, Olfaction.M,
         Hearing.M, Interoception.M, Head.M, Mouth_throat.M, Hand_arm.M,
         Torso.M, Foot_leg.M)

cors = cor(cols)

```


What about tokenization? (These might all be cases where the disambiguating token is multiple tokens.)


```{r}

### Load data
directory_path <- "../../data/processed/rawc/pythia/token_analysis/"
csv_files <- list.files(path = directory_path, pattern = "*.csv", full.names = TRUE)
csv_list <- csv_files %>%
  map(~ read_csv(.))
df_pythia_token_analysis <- bind_rows(csv_list) %>%
  mutate(Layer = Layer + 1,
         Head = Head + 1)

nrow(df_pythia_token_analysis)


### Final step
df_pythia_token_analysis_final = df_pythia_token_analysis %>%
  filter(step == 143000) %>%
  filter(Head == 2) %>%
  filter(Layer == 3) %>%
  rename(Attention_new = Attention,
         Entropy_new = Entropy) %>%
  select(Head, Layer, Attention_new, Entropy_new, attention_to_previous,
         step, word, sentence, 
         num_target_tokens, num_disambiguating_tokens)


### Merge
df_merged_h2 = just_h2 %>%
  select(Head, Layer, Entropy, Attention, step, word, sentence, disambiguating_word,
         Attention_to_disambiguating, attention_diff) %>%
  inner_join(df_pythia_token_analysis_final) %>%
  mutate(attention_diff2 = attention_to_previous - Attention)
nrow(df_merged_h2)


### Compare to original attention_diff
df_merged_h2 %>%
  ggplot(aes(x = attention_diff,
             y = attention_diff2,
             color = num_disambiguating_tokens)) +
  geom_point(size = 2, alpha = .6) +
  theme_minimal() +
  theme(text = element_text(size = 15),
        legend.position = "bottom") +
  labs(x = "Original Difference (L2, H3)",
       y = "Recalculated Difference (L2, H3)",
       color = "#Tokens in Disambiguating Word") + 
  scale_color_viridis(option = "mako")

### Calculate in model
m = lm(data = df_merged_h2, attention_diff ~ num_disambiguating_tokens)
summary(m)


### Comparison for l3, h2 by step
df_merged_h2 %>%
  select(attention_diff, attention_diff2) %>%
  pivot_longer(cols = c(attention_diff, attention_diff2),
               names_to = "Comparison", values_to = "Difference") %>%
  ggplot(aes(x = Difference, 
             fill = Comparison)) +
  geom_density(size = 0, alpha = .6) +
  theme_minimal() +
    geom_vline(xintercept = 0, linetype = "dotted", size =1.5) +
  labs(x = "Attention Difference (L3, H2)",
       fill = "") +
  scale_fill_viridis_d(option = "mako", 
                       labels = c("Entire Word", "Final Token")) +
  theme(text = element_text(size = 15),
        legend.position = "bottom")


### Summaries
summary(df_merged_h2$attention_diff)
sd(df_merged_h2$attention_diff)
summary(df_merged_h2$attention_diff2)
sd(df_merged_h2$attention_diff2)

```


