---
title: "Analysis of HuggingFace Models Attention"
author: "Sean Trott and Pam Rivière"
date: "November 20, 2024"
output:
  html_document:
    keep_md: yes
    toc: yes
    toc_float: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(dpi = 300, fig.format = "pdf")
```


```{r include=FALSE}
library(tidyverse)
library(lmtest)
library(forcats)
library(broom)
library(lme4)
library(viridis)
library(ggridges)
library(lmerTest)
library(ggcorrplot)

all_colors <- viridis::viridis(10, option = "mako")
my_colors <- all_colors[c(3, 5, 7)]  # Selecting specific colors from the palette
```

# Load Pythia data


```{r}
# setwd("/Users/seantrott/Dropbox/UCSD/Research/Ambiguity/SSD/entangled_meanings/src/analysis/")
directory_path <- "../../data/processed/rawc/pythia/attention/"
csv_files <- list.files(path = directory_path, pattern = "*.csv", full.names = TRUE)
csv_list <- csv_files %>%
  map(~ read_csv(.))
df_pythia_models <- bind_rows(csv_list) %>%
  mutate(Layer = Layer + 1)

table(df_pythia_models$mpath)

```

# Entropy over time

```{r}
df_entropy_over_time = df_pythia_models %>% 
  group_by(mpath, step, revision, Layer, n_params) %>%
  summarise(mean_entropy = mean(Entropy)) %>%
  mutate(step_modded = step + 1) %>%
  group_by(mpath, Layer) %>%
  mutate(
    entropy_diff = mean_entropy - lag(mean_entropy)
  )

df_entropy_over_time %>%
  ggplot(aes(x = step_modded,
             y = mean_entropy,
             color = factor(Layer))) +
  geom_point(size = 4, alpha = .7) +
  geom_line(size = 2, alpha = .7) +
  theme_minimal() +
  labs(x = "Training Step (Log10)",
       y = "Entropy Across Heads",
       color = "Layer") +
  scale_x_log10() +
  theme(text = element_text(size = 15),
        legend.position = "bottom") +
  scale_color_viridis(option = "mako", discrete=TRUE) +
  facet_wrap(~reorder(mpath, n_params))

df_entropy_over_time %>%
  ggplot(aes(x = step_modded,
             y = entropy_diff,
             color = factor(Layer))) +
  geom_point(size = 4, alpha = .7) +
  geom_line(size = 2, alpha = .7) +
  theme_minimal() +
  geom_hline(yintercept = 0, linetype = "dotted", size = 1) +
  labs(x = "Training Step (Log10)",
       y = "Entropy Delta",
       color = "") +
  scale_x_log10() +
  theme(text = element_text(size = 15),
        legend.position = "bottom") +
  scale_color_viridis(option = "mako", discrete=TRUE) +
  facet_wrap(~mpath)
```

# Attention at final timestep

```{r}
df_attention_final_step_prop_layer = df_pythia_models %>%
  filter(step == 143000) %>%
  group_by(mpath, step, Head, Layer, n_params) %>%
  summarise(mean_attention = mean(Attention)) %>%
  group_by(mpath, Layer, n_params) %>%
  summarise(max_attention_head = max(mean_attention)) %>%
  group_by(mpath) %>%
  mutate(max_layer = max(Layer),
         prop_layer = Layer / max_layer) 

df_attention_final_step_prop_layer %>%
  mutate(binned_prop_layer = ntile(prop_layer, 6)) %>%
  mutate(prop_binned = binned_prop_layer / 6) %>%
  ggplot(aes(x = prop_binned,
             y = max_attention_head)) +
  stat_summary(
    aes(group = mpath,
        color = mpath),  
    fun = mean,    
    geom = "line",        
    size = 2              
  ) +
  stat_summary(
    aes(group = mpath, 
        fill = mpath), 
    fun.data = mean_se,    
    geom = "ribbon",  
    alpha = 0.2,   
    color = NA     
  ) +
  theme_minimal() +
  labs(x = "Layer Depth Ratio",
       y = "Attention to Disambiguating Word") +
  scale_color_viridis(option = "mako", discrete = TRUE) +
  theme(text = element_text(size = 15)) 
```

# Attention over time

## Averaging across heads, per layer


**TODO**: instead, maybe get "max head" per layer?

**TODO**: Do this also as a function 

```{r}
df_attention_over_time = df_pythia_models %>% 
  group_by(mpath, step, revision, Head, Layer, n_params) %>%
  summarise(mean_attention = mean(Attention)) %>%
  group_by(mpath, step, revision, Layer, n_params) %>%
  summarise(max_attention_head = max(mean_attention)) %>%
  mutate(step_modded = step + 1) %>%
  group_by(mpath, Layer) %>%
  mutate(
    attention_diff = max_attention_head - lag(max_attention_head)
  )

df_attention_over_time %>%
  ggplot(aes(x = step_modded,
             y = max_attention_head,
             color = factor(Layer))) +
  geom_point(size = 4, alpha = .7) +
  geom_line(size = 2, alpha = .7) +
  theme_minimal() +
  geom_vline(xintercept = 1000, 
             linetype = "dotted", 
             size = 1.2) +
  labs(x = "Training Step (Log10)",
       y = "Attention to Disambiguating Word",
       color = "Layer") +
  scale_x_log10() +
  theme(text = element_text(size = 15),
        legend.position = "bottom") +
  scale_color_viridis(option = "mako", discrete=TRUE) +
  facet_wrap(~reorder(mpath, n_params))

df_attention_over_time %>%
  ggplot(aes(x = step_modded,
             y = attention_diff,
             color = factor(Layer))) +
  geom_point(size = 4, alpha = .7) +
  geom_line(size = 2, alpha = .7) +
  theme_minimal() +
  geom_hline(yintercept = 0, linetype = "dotted", size = 1) +
  labs(x = "Training Step (Log10)",
       y = "Attention Delta",
       color = "") +
  scale_x_log10() +
  theme(text = element_text(size = 15),
        legend.position = "bottom") +
  scale_color_viridis(option = "mako", discrete=TRUE) +
  facet_wrap(~mpath)
```



## Attention by head


**TODO**: calculate attention differences by head
```{r}
df_attention_over_time = df_pythia_models %>% 
  # filter(Layer == 3) %>%
  group_by(mpath, revision, Layer, Head, step, n_params) %>%
  summarise(mean_attention = mean(Attention)) %>%
  mutate(step_modded = step + 1) %>%
  group_by(mpath, step, Head) %>%
  mutate(
    attention_diff = mean_attention - lag(mean_attention),
    Head = Head + 1
  )

df_attention_over_time %>%
  # filter(Layer == 3) %>%
  filter(Layer %in% c(3, 4, 5, 6)) %>%
  ggplot(aes(x = step_modded,
             y = mean_attention,
             color = factor(Head))) +
  geom_point(size = 4, alpha = .7) +
  geom_line(size = 2, alpha = .7) +
  theme_minimal() +
  labs(x = "Training Step (Log10)",
       y = "Attention to Disambiguating Word",
       color = "Attention Head") +
  scale_x_log10() +
  theme(text = element_text(size = 15),
        legend.position = "bottom") +
  scale_color_viridis(option = "mako", discrete=TRUE) +
  facet_wrap(~reorder(mpath, n_params) + Layer, ncol = 4)

```


Testing out Pam's idea:

```{r}
df_attention_over_time %>%
  filter(mpath == "EleutherAI/pythia-14m") %>%
  ggplot(aes(x = factor(step, levels = sort(unique(step))), 
             y = factor(Head), # Use only "Head" as y-axis
             fill = mean_attention)) +
  geom_tile() +
  facet_grid(Layer ~ ., scales = "free_y", space = "free") + # Separate by Layer
  theme_minimal() +
  scale_x_discrete(
    breaks = sort(unique(df_attention_over_time$step))[seq(1, length(unique(df_attention_over_time$step)), 
                                      by = 15)] # Every 10th step
  ) +
  labs(x = "Step",
       y = "Head",
       fill = "Attention to Disambiguating Word") +
  theme(text = element_text(size = 15),
        strip.text.y = element_text(angle = 0), # Keep facet labels horizontal
        axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "bottom")

df_attention_over_time %>%
  filter(mpath == "EleutherAI/pythia-14m") %>%
  ggplot(aes(x = step,  # Keep step as continuous
             y = factor(Head),  # Combine Layer and Head hierarchically
             fill = mean_attention)) +
  geom_tile(width = 1) +
  theme_minimal() +
  facet_grid(Layer ~ ., scales = "free_y", space = "free") + # Separate by Layer
  labs(x = "Step",
       y = "Layer_Head",
       fill = "Attention to Disambiguating Word") +
  theme(text = element_text(size = 15),
        axis.text.x = element_text(angle = 90, hjust = 1))
```


### Attention at sentence-level

```{r}
df_pythia_models %>%
  filter(mpath == "EleutherAI/pythia-14m") %>%
  filter(Layer == 3) %>%
  filter(step %in% c(512, 1000, 2000, 143000)) %>%
  ggplot(aes(x = Attention,
             y = factor(Head),
             fill = factor(mpath))) +
  geom_density_ridges2(aes(height = ..density..), 
                       color=gray(0.25), 
                       alpha = .7, 
                       scale=.85, 
                       # size=1, 
                       size = 0,
                       stat="density") +
  labs(x = "Attention to Disambiguating Word",
       y = "Attention Head",
       fill = "") +
  theme_minimal() +
  scale_fill_viridis(option = "mako", discrete=TRUE) +
  theme(text = element_text(size = 15),
        legend.position="none") +
  facet_wrap(~reorder(revision, step))
```


## Overlaying attention and R2

**TODO**: Just messing around with this part, doesn't actaully run without loading the distance data.

**TODO**: Need to also get layer information for $R^2$.

```{r}
df_merged_r2_attention = df_attention_over_time %>%
  mutate(Layer = Layer + 1) %>%
  # select(-Layer) %>%
  inner_join(df_best_r2)

scale_factor <- max(df_merged_r2_attention$mean_attention, na.rm = TRUE) / 
                max(df_merged_r2_attention$r2, na.rm = TRUE)

df_merged_r2_attention %>%
  ggplot(aes(x = step_modded)) +
  # Primary Y-axis: Attention
  geom_line(aes(y = mean_attention, color = factor(Layer)), size = 2, alpha = .7) +
  # Secondary Y-axis: R-squared
  geom_line(aes(y = r2 * scale_factor, color = factor(Layer)), 
            size = 2, linetype = "dotted") +
  scale_y_continuous(
    name = "Attention to Disambiguating Word",
    sec.axis = sec_axis(~ . / scale_factor, name = "R-squared")
  ) +
  scale_x_log10() +
  theme_minimal() +
  labs(
    x = "Training Step (Log10)",
    color = ""
  ) +
  facet_wrap(~Layer) +
  scale_color_viridis(option = "mako", discrete=TRUE) +
  theme(
    text = element_text(size = 15),
    legend.position = "bottom"
  )
```



