---
title: "Analysis of HuggingFace Models Attention"
author: "Sean Trott and Pam Rivière"
date: "November 20, 2024"
output:
  html_document:
    keep_md: yes
    toc: yes
    toc_float: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(dpi = 300, fig.format = "pdf")
```


```{r include=FALSE}
library(tidyverse)
library(lmtest)
library(forcats)
library(broom)
library(lme4)
library(viridis)
library(ggridges)
library(lmerTest)
library(ggrepel)

all_colors <- viridis::viridis(10, option = "mako")
my_colors <- all_colors[c(3, 5, 7)]  # Selecting specific colors from the palette
```

# Load Pythia data


```{r}
# setwd("/Users/seantrott/Dropbox/UCSD/Research/Ambiguity/SSD/entangled_meanings/src/analysis/")
directory_path <- "../../data/processed/rawc/pythia/attention/"
csv_files <- list.files(path = directory_path, pattern = "*.csv", full.names = TRUE)
csv_list <- csv_files %>%
  map(~ read_csv(.))
df_pythia_models <- bind_rows(csv_list) %>%
  mutate(Layer = Layer + 1)

table(df_pythia_models$mpath)

```

# Entropy over time

For this developmental analysis, we focus on Pythia-14m.


## Mean entropy


```{r}
### By model
df_pythia_models %>%
  filter(step == 143000) %>%
  group_by(mpath) %>%
  summarise(mean_entropy = mean(Entropy, na.rm = TRUE),
            sd_entropy = sd(Entropy, na.rm = TRUE),
            count = n())


df_final = df_pythia_models %>%
  filter(step == 143000) %>%
  group_by(mpath, n_params, Layer, Head) %>%
  summarise(mean_entropy = mean(Entropy, na.rm = TRUE),
            sd_entropy = sd(Entropy, na.rm = TRUE),
            count = n()) %>%
  group_by(mpath) %>%
  mutate(mean_entropy_scaled = scale(mean_entropy)) 
mod_full = lmer(data = df_final, mean_entropy_scaled ~ Layer + 
                  (1 | mpath) + (1 | Head))

summary(mod_full)


df_final %>%
  group_by(mpath) %>%
  mutate(mean_entropy_scaled = scale(mean_entropy)) %>%
  ggplot(aes(x = Layer,
             y = Head,
             fill = mean_entropy_scaled)) +
  geom_tile() +
  labs(x = "Layer",
       y = "Head",
       fill = "Mean Entropy (Z-scored)") +
  scale_fill_gradient2(low = "blue",
                       mid = "white",
                       high = "red",
                       # midpoint = 0, 
                       name = "Mean Entropy (Z-scored)") +
  theme_minimal() +
  theme(text = element_text(size = 15),
        strip.text.y = element_text(angle = 0), # Keep facet labels horizontal
        axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "bottom") +
  facet_wrap(~reorder(mpath, n_params))

```


## Mean entropy for 14M

```{r}
summary_df <- df_pythia_models %>%
  filter(mpath == "EleutherAI/pythia-14m") %>%
  group_by(step) %>%
  summarise(
    mean_entropy = mean(Entropy, na.rm = TRUE),
    se_entropy = sd(Entropy, na.rm = TRUE) / sqrt(n()),
    .groups = "drop"
  )


summary_df %>%
  filter(step %in% c(1, 512, 143000))

ggplot(summary_df, aes(x = step, y = mean_entropy, color = mpath)) +
  geom_line(color = "black") +  # Lineplot for mean entropy
  labs(
    title = "",
    x = "Step",
    y = "Mean Entropy"
  ) +
  theme_minimal() +
  scale_x_log10() +
  scale_y_continuous(limits = c(1.72, 1.85)) +
  theme(text = element_text(size = 15),
        legend.position = "bottom") +
  geom_vline(xintercept = 1000, 
             linetype = "dotted", 
             size = 1.2)



pythia_14m = df_pythia_models %>%
  filter(mpath == "EleutherAI/pythia-14m") %>%
  mutate(step_modded = step + 1) %>%
  group_by(step_modded, Head, Layer) %>%
  summarise(avg_entropy = mean(Entropy))
nrow(pythia_14m)

mean(pythia_14m$avg_entropy)
sd(pythia_14m$avg_entropy)
summary(pythia_14m$avg_entropy)


summary(lmer(data = pythia_14m, 
             avg_entropy ~ log10(step_modded) + (1 | Head) + (1 | Layer)))



summary(lm(data = pythia_14m, 
             avg_entropy ~ log10(step_modded) * Layer))

```




# Attention at final step

```{r}
df_final_step = df_pythia_models %>%
  filter(step == 143000)
nrow(df_final_step)

df_attn_by_head = df_final_step %>%
  mutate(Head = Head + 1) %>%
  group_by(mpath, Layer, Head, n_params) %>%
  summarise(mean_attention = mean(Attention)) %>%
  group_by(mpath) %>%
  mutate(max_layer = max(Layer),
         prop_layer = Layer / max_layer) %>%
  mutate(binned_prop_layer = ntile(prop_layer, 6)) %>%
  mutate(prop_binned = binned_prop_layer / 6) 


df_attn_by_head %>%
  ggplot(aes(x = Layer,
             y = Head,
             fill = mean_attention)) +
  geom_tile() +
  labs(x = "Layer",
       y = "Head",
       fill = "Attention to Disambiguating Word") +
  scale_fill_gradient2(low = "blue",
                       mid = "white",
                       high = "red",
                       midpoint = 0, 
                       name = "Attention to Disambiguating Word") +
  theme_minimal() +
  theme(text = element_text(size = 15),
        strip.text.y = element_text(angle = 0), # Keep facet labels horizontal
        axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "bottom") +
  facet_wrap(~reorder(mpath, n_params))

df_attn_by_head %>%
  ggplot(aes(x = Layer,
             y = Head,
             fill = mean_attention)) +
  geom_tile() +
  labs(x = "Layer",
       y = "Head",
       fill = "Attention to Disambiguating Word") +
  scale_fill_gradient2(low = "blue",
                       mid = "white",
                       high = "red",
                       midpoint = 0, 
                       name = "Attention to Disambiguating Word") +
  theme_minimal() +
  theme(text = element_text(size = 15),
        strip.text.y = element_text(angle = 0), # Keep facet labels horizontal
        axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "bottom") +
  facet_wrap(~reorder(mpath, n_params),
             scales = "free")



df_attn_by_head %>%
  group_by(mpath, prop_binned, n_params) %>%
  summarise(mean_attention = mean(mean_attention)) %>%
  ggplot(aes(x = prop_binned,
             y = reorder(mpath, n_params),
             fill = mean_attention)) +
  geom_tile() +
  labs(x = "Layer Depth Ratio",
       y = "",
       fill = "Attention to Disambiguating Word") +
  scale_fill_gradient2(low = "blue",
                       mid = "white",
                       high = "red",
                       midpoint = 0, 
                       name = "Attention to Disambiguating Word") +
  theme_minimal() +
  theme(text = element_text(size = 15),
        strip.text.y = element_text(angle = 0), # Keep facet labels horizontal
        axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "bottom") 



df_attn_by_head %>%
  filter(mpath == "EleutherAI/pythia-14m") %>%
  ggplot(aes(x = Layer,
             y = Head,
             fill = mean_attention)) +
  geom_tile() +
  labs(x = "Layer",
       y = "Head",
       fill = "Attention to Disambiguating Word") +
  scale_fill_gradient2(low = "blue",
                       mid = "white",
                       high = "red",
                       midpoint = 0, 
                       name = "Attention to Disambiguating Word") +
  theme_minimal() +
  theme(text = element_text(size = 15),
        strip.text.y = element_text(angle = 0), # Keep facet labels horizontal
        axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "bottom")






final_step_14m = df_attn_by_head %>%
  filter(mpath == "EleutherAI/pythia-14m") 
summary(final_step_14m$mean_attention)
sd(final_step_14m$mean_attention)

final_step_14m %>%
  ggplot(aes(x = mean_attention)) +
  geom_histogram(alpha = .6, bins = 10) +
  theme_minimal() +
  labs(x = "Attention to Disambiguating Word") +
  theme(text = element_text(size = 15),
        legend.position = "bottom")


final_step_14m %>%
  mutate(mean_attention_scaled = scale(mean_attention)) %>%
  ggplot(aes(x = mean_attention_scaled)) +
  geom_histogram(alpha = .6, bins = 10) +
  theme_minimal() +
  labs(x = "Z-scored Attention to Disambiguating Word") +
  theme(text = element_text(size = 15),
        legend.position = "bottom")


df_attn_by_head %>%
  group_by(mpath) %>%
  mutate(mean_attention_scaled = scale(mean_attention)) %>%
  ggplot(aes(x = mean_attention_scaled,
             y = ..density..)) +
  geom_histogram(alpha = .6, bins = 10) +
  theme_minimal() +
  labs(x = "Z-scored Attention to Disambiguating Word") +
  theme(text = element_text(size = 15),
        legend.position = "bottom") +
  facet_wrap(~reorder(mpath, n_params))


### Very attentive heads
df_most_attentive_heads = df_attn_by_head %>%
  group_by(mpath) %>%
  mutate(mean_attention_scaled = scale(mean_attention),
         num_heads = max(Layer) * max(Head)) %>%
  filter(mean_attention_scaled > 3) 

df_most_attentive_heads %>%
  group_by(mpath) %>%
  summarise(count = n(),
            num_heads = max(num_heads),
            proportion = count/ num_heads)
  


```



# Main analysis: attention over time for 14m

## Averaging across heads, per layer

```{r}
df_attention_over_time = df_pythia_models %>% 
  filter(mpath %in% c("EleutherAI/pythia-14m")) %>%
  group_by(mpath, step, revision, Head, Layer, n_params) %>%
  summarise(mean_attention = mean(Attention)) %>%
  group_by(mpath, step, revision, Layer, n_params) %>%
  summarise(max_attention_head = max(mean_attention)) %>%
  mutate(step_modded = step + 1) %>%
  group_by(mpath, Layer) %>%
  mutate(
    attention_diff = max_attention_head - lag(max_attention_head)
  )

df_attention_over_time %>%
  ggplot(aes(x = step_modded,
             y = max_attention_head,
             color = factor(Layer))) +
  geom_point(size = 4, alpha = .7) +
  geom_line(size = 2, alpha = .7) +
  theme_minimal() +
  geom_vline(xintercept = 1000, 
             linetype = "dotted", 
             size = 1.2) +
  labs(x = "Training Step (Log10)",
       y = "Attention to Disambiguating Word",
       color = "Layer") +
  scale_x_log10() +
  geom_vline(xintercept = 1000, 
             linetype = "dotted", 
             size = 1.2) +
  theme(text = element_text(size = 15),
        legend.position = "bottom") +
  scale_color_viridis(option = "mako", discrete=TRUE) +
  facet_wrap(~reorder(mpath, n_params))


```



## Attention by head

```{r}
df_attention_over_time = df_pythia_models %>% 
  filter(mpath %in% c("EleutherAI/pythia-14m")) %>%
  # filter(Layer == 3) %>%
  group_by(mpath, revision, Layer, Head, step, n_params) %>%
  summarise(mean_attention = mean(Attention)) %>%
  mutate(step_modded = step + 1) %>%
  group_by(mpath, step, Head) %>%
  mutate(
    attention_diff = mean_attention - lag(mean_attention),
    Head = Head + 1
  )

df_attention_over_time %>%
  # filter(Layer == 3) %>%
  ggplot(aes(x = step_modded,
             y = mean_attention,
             color = factor(Head))) +
  geom_point(size = 4, alpha = .7) +
  geom_line(size = 2, alpha = .7) +
  theme_minimal() +
  labs(x = "Training Step (Log10)",
       y = "Attention to Disambiguating Word",
       color = "Attention Head") +
  scale_x_log10() +
    geom_vline(xintercept = 1000, 
             linetype = "dotted", 
             size = 1.2) +
  theme(text = element_text(size = 15),
        legend.position = "bottom") +
  scale_color_viridis(option = "mako", discrete=TRUE) +
  facet_wrap(~Layer)

df_attention_over_time %>%
  filter(Layer == 3) %>%
  # filter(Head %in% c(1, 2)) %>%
  ggplot(aes(x = step_modded,
             y = mean_attention,
             color = factor(Head))) +
  geom_point(size = 4, alpha = .7) +
  geom_line(size = 2, alpha = .7) +
  theme_minimal() +
  labs(x = "Training Step (Log10)",
       y = "Attention to Disambiguating Word",
       color = "Attention Head (Layer 3)") +
  scale_x_log10() +
    geom_vline(xintercept = 1000, 
             linetype = "dotted", 
             size = 1.2) +
  theme(text = element_text(size = 15),
        legend.position = "bottom") +
  scale_color_viridis(option = "mako", discrete=TRUE) 




```





### Attention at sentence-level

```{r}
df_pythia_models %>%
  filter(mpath == "EleutherAI/pythia-14m") %>%
  filter(Layer == 3) %>%
  filter(step %in% c(512, 1000, 2000, 143000)) %>%
  ggplot(aes(x = Attention,
             y = factor(Head),
             fill = factor(mpath))) +
  geom_density_ridges2(aes(height = ..density..), 
                       color=gray(0.25), 
                       alpha = .7, 
                       scale=.85, 
                       # size=1, 
                       size = 0,
                       stat="density") +
  labs(x = "Attention to Disambiguating Word",
       y = "Attention Head",
       fill = "") +
  theme_minimal() +
  scale_fill_viridis(option = "mako", discrete=TRUE) +
  theme(text = element_text(size = 15),
        legend.position="none") +
  facet_wrap(~reorder(revision, step))
```


# Attention vs. R2


## Load R2 data

```{r}
# setwd("/Users/seantrott/Dropbox/UCSD/Research/Ambiguity/SSD/entangled_meanings/src/analysis/")
directory_path <- "../../data/processed/rawc/pythia/distances/"
csv_files <- list.files(path = directory_path, pattern = "*.csv", full.names = TRUE)
csv_list <- csv_files %>%
  map(~ read_csv(.))
df_pythia_models <- bind_rows(csv_list)

table(df_pythia_models$mpath)

df_best_r2 = df_pythia_models %>% 
  filter(mpath %in% c("EleutherAI/pythia-14m")) %>%
  group_by(mpath, revision, step, Layer, n_params) %>%
  summarise(r = cor(Distance, mean_relatedness, method = "pearson"),
            r2 = r ** 2) %>%
  group_by(mpath, step, n_params) %>%
  mutate(step_modded = step + 1) 
```

## Merge

```{r}
df_merged_r2_attention = df_attention_over_time %>%
  inner_join(df_best_r2)


scale_factor <- max(df_merged_r2_attention$mean_attention, na.rm = TRUE) / 
                max(df_merged_r2_attention$r2, na.rm = TRUE)

df_merged_r2_attention %>%
  filter(Head == 2) %>%
  filter(Layer == 3) %>%
  ggplot(aes(x = step_modded)) +
  # Primary Y-axis: Attention
  geom_line(aes(y = mean_attention), size = 2, alpha = .7) +
  # Secondary Y-axis: R-squared
  geom_line(aes(y = r2 * scale_factor), 
            size = 2, linetype = "dotted") +
  scale_y_continuous(
    name = "Attention to Disambiguating Word",
    sec.axis = sec_axis(~ . / scale_factor, name = "R-squared")
  ) +
  scale_x_log10() +
  theme_minimal() +
  labs(
    x = "Training Step (Log10)",
    color = ""
  ) +
  # facet_wrap(~Head) +
  scale_color_viridis(option = "mako", discrete=TRUE) +
  theme(
    text = element_text(size = 15),
    legend.position = "bottom"
  )

```


## Analysis of attention and r2

```{r}
### Make actual best R2
df_best_r2_actual = df_best_r2 %>%
  group_by(revision) %>%
  slice_max(r2) %>%
  select(revision, step, r2)

### 
df_merged_r2_attention_actual = df_attention_over_time %>%
  inner_join(df_best_r2_actual)

# Pivot the dataframe to wider format
df_wide <- df_merged_r2_attention_actual %>%
  select(revision, Layer, Head, step, mean_attention, r2) %>%
  group_by(revision) %>%
  slice_max(r2) %>%
  pivot_wider(names_from = c(Layer, Head), 
              values_from = c(mean_attention), names_sep = "_") %>%
  ungroup()

### Target variable
y <- df_wide$r2

# Run separate regressions for each predictor and measure R^2 and coefficient
regression_results <- lapply(names(df_wide)[!(names(df_wide) %in% c("mpath", "revision", "step", "n_params", "step_modded", "r", "r2"))], function(var) {
  model <- lm(y ~ df_wide[[var]])
  summary_model <- summary(model)
  p_value <- coef(summary_model)[2, "Pr(>|t|)"]
  list(Feature = var, 
       R2 = summary_model$r.squared, 
       Coefficient = coef(model)[2],
       P_Value = p_value)
})


regression_results_df <- do.call(rbind, regression_results) %>% as.data.frame()

# Coerce types and apply multiple comparisons correction
regression_results_df <- regression_results_df %>%
  mutate(R2 = as.numeric(R2),
         Coefficient = as.numeric(Coefficient),
         P_Value = as.numeric(P_Value),
         P_Value_adj = p.adjust(P_Value, method = "fdr"))


regression_results_df %>%
  filter(Coefficient > 0) %>%
  filter(P_Value_adj < .05)
```



# Replication for 410m


## Averaging across heads, per layer

```{r}
df_attention_over_time = df_pythia_models %>% 
  filter(mpath %in% c("EleutherAI/pythia-410m")) %>%
  group_by(mpath, step, revision, Head, Layer, n_params) %>%
  summarise(mean_attention = mean(Attention)) %>%
  group_by(mpath, step, revision, Layer, n_params) %>%
  summarise(max_attention_head = max(mean_attention)) %>%
  mutate(step_modded = step + 1) %>%
  group_by(mpath, Layer) %>%
  mutate(
    attention_diff = max_attention_head - lag(max_attention_head)
  )

df_attention_over_time %>%
  ggplot(aes(x = step_modded,
             y = max_attention_head,
             color = factor(Layer))) +
  geom_point(size = 4, alpha = .7) +
  geom_line(size = 2, alpha = .7) +
  theme_minimal() +
  geom_vline(xintercept = 1000, 
             linetype = "dotted", 
             size = 1.2) +
  labs(x = "Training Step (Log10)",
       y = "Max. Attention to Disambiguating Word",
       color = "Layer") +
  scale_x_log10() +
  geom_vline(xintercept = 1000, 
             linetype = "dotted", 
             size = 1.2) +
  theme(text = element_text(size = 15),
        legend.position = "bottom") +
  scale_color_viridis(option = "mako", discrete=TRUE) +
  facet_wrap(~Layer)


```



## Attention by head

```{r}
df_attention_over_time = df_pythia_models %>% 
  filter(mpath %in% c("EleutherAI/pythia-410m")) %>%
  # filter(Layer == 3) %>%
  group_by(mpath, revision, Layer, Head, step, n_params) %>%
  summarise(mean_attention = mean(Attention)) %>%
  mutate(step_modded = step + 1) %>%
  group_by(mpath, step, Head) %>%
  mutate(
    attention_diff = mean_attention - lag(mean_attention),
    Head = Head + 1
  )

df_attention_over_time %>%
  # filter(Layer == 3) %>%
  ggplot(aes(x = step_modded,
             y = mean_attention,
             color = factor(Head))) +
  geom_point(size = 4, alpha = .7) +
  geom_line(size = 2, alpha = .7) +
  theme_minimal() +
  labs(x = "Training Step (Log10)",
       y = "Attention to Disambiguating Word",
       color = "Attention Head") +
  scale_x_log10() +
    geom_vline(xintercept = 1000, 
             linetype = "dotted", 
             size = 1.2) +
  theme(text = element_text(size = 15),
        legend.position = "bottom") +
  scale_color_viridis(option = "mako", discrete=TRUE) +
  facet_wrap(~Layer)

df_attention_over_time %>%
  filter(Layer %in% c(1, 4, 5, 6, 10, 22, 24)) %>%
  # filter(Head %in% c(1, 2)) %>%    
  ggplot(aes(x = step_modded,
             y = mean_attention,
             color = factor(Head))) +
  geom_point(size = 4, alpha = .7) +
  geom_line(size = 2, alpha = .7) +
  theme_minimal() +
  labs(x = "Training Step (Log10)",
       y = "Attention to Disambiguating Word",
       color = "Attention Head (Layer 3)") +
  scale_x_log10() +
    geom_vline(xintercept = 1000, 
             linetype = "dotted", 
             size = 1.2) +
  theme(text = element_text(size = 15),
        legend.position = "bottom") +
  scale_color_viridis(option = "mako", discrete=TRUE) +
  facet_wrap(~Layer)




```


## 1-back for 410m

### Over time

```{r}
df_attention_over_time = df_pythia_models %>% 
  filter(mpath %in% c("EleutherAI/pythia-410m")) %>%
  group_by(mpath, revision, Layer, Head, step, n_params) %>%
  mutate(attention_diff = Attention - Attention_1back,
         attention_ratio = Attention / Attention_1back) %>%
  summarise(mean_attention_diff = mean(attention_diff),
            mean_attention_ratio = mean(attention_ratio),
            se_diff = sd(attention_diff) / n()) %>%
  mutate(step_modded = step + 1)

df_attention_over_time %>%
  ggplot(aes(x = step_modded,
             y = mean_attention_diff,
             color = factor(Head))) +
  geom_point(size = 2, alpha = .7) +
  geom_line(size = 2, alpha = .7) +
  theme_minimal() +
  labs(x = "Training Step (Log10)",
       y = "Attention Difference",
       color = "Attention Head") +
  scale_x_log10() +
    geom_vline(xintercept = 1000, 
             linetype = "dotted", 
             size = 1.2) +
    geom_hline(yintercept = 0, 
             linetype = "dashed", 
             size = 1.2,
             color = "black") +
  theme(text = element_text(size = 15),
        legend.position = "bottom") +
  scale_color_viridis(option = "mako", discrete=TRUE) +
  facet_wrap(~Layer)

df_attention_over_time %>%
  filter(Layer %in% c(1, 4, 5, 6, 10, 22, 24)) %>%
  ggplot(aes(x = step_modded,
             y = mean_attention_diff,
             color = factor(Head))) +
  geom_point(size = 2, alpha = .7) +
  geom_line(size = 2, alpha = .7) +
  theme_minimal() +
  labs(x = "Training Step (Log10)",
       y = "Attention Difference",
       color = "Attention Head") +
  scale_x_log10() +
    geom_vline(xintercept = 1000, 
             linetype = "dotted", 
             size = 1.2) +
    geom_hline(yintercept = 0, 
             linetype = "dashed", 
             size = 1.2,
             color = "black") +
  theme(text = element_text(size = 15),
        legend.position = "bottom") +
  scale_color_viridis(option = "mako", discrete=TRUE) +
  facet_wrap(~Layer)

df_attention_over_time %>%
  ggplot(aes(x = step_modded,
             y = mean_attention_ratio,
             color = factor(Head))) +
  geom_point(size = 2, alpha = .7) +
  geom_line(size = 2, alpha = .7) +
  theme_minimal() +
  labs(x = "Training Step (Log10)",
       y = "Attention Ratio",
       color = "Attention Head") +
  scale_x_log10() +
    geom_vline(xintercept = 1000, 
             linetype = "dotted", 
             size = 1.2) +
    geom_hline(yintercept = 1, 
             linetype = "dashed", 
             size = 1.2,
             color = "black") +
  theme(text = element_text(size = 15),
        legend.position = "bottom") +
  scale_color_viridis(option = "mako", discrete=TRUE) +
  facet_wrap(~Layer)
```


## T-test for subtraction analysis

```{r}
t_test_results <- df_pythia_models %>%
  filter(mpath == "EleutherAI/pythia-410m") %>%
  group_by(mpath, Layer, Head, n_params, step) %>%
  summarise(
    t_test = list(
      t.test(
        Attention,
        Attention_1back,
        paired = TRUE,
        alternative = "greater"  # One-tailed test for greater values
      )
    ),
    .groups = "drop"
  ) %>%
  mutate(
    p_value = map_dbl(t_test, ~ .x$p.value),          # Extract p-value
    t_statistic = map_dbl(t_test, ~ .x$statistic)    # Extract t-statistic
  )


# Apply multiple testing correction
t_test_results <- t_test_results %>%
  group_by(mpath) %>%
  mutate(
    p_adjusted = p.adjust(p_value, method = "fdr")  # Adjust p-values (FDR)
  ) %>%
  mutate(p_adj2 = case_when(
    p_adjusted < .05 ~ p_adjusted,
    p_adjusted >= .05 ~ 1
  ))

# Filter significant results
significant_results <- t_test_results %>%
  filter(p_adjusted < 0.05) 


### Significant at final step
significant_results %>%
  filter(step == 143000) %>%
  ggplot(aes(x = Layer, y = Head, fill = -log10(p_adjusted))) +
  geom_tile() +
  scale_fill_gradient(low = "white", high = "red", name = "-log10(FDR p-value)") +
  theme_minimal() +
  labs(
    title = "Significant Heads/Layers",
    x = "Layer",
    y = "Head"
  )

significant_results %>%
  filter(step == 143000) %>%
  select(Layer, Head, t_statistic, p_adj2)



### How many timesteps?
significant_results %>%
  filter(mpath == "EleutherAI/pythia-410m") %>%
  group_by(Layer, Head) %>%
  summarise(count = n(),
            prop = count / 20)

### Geom rug
df_attention_over_time %>%
  filter(mpath == "EleutherAI/pythia-410m") %>%
  inner_join(t_test_results) %>%
  filter(Layer == 1) %>%
  mutate(is_significant = p_adjusted < .05,
         Head = factor(Head)) %>%
  ggplot(aes(x = step_modded,
             y = mean_attention_diff,
             color = Head)) +
  geom_line(size = 1.2, alpha = 0.8) +
  geom_point(alpha = 0.6) +
  # Head-specific rug layer
  geom_rug(data = ~ .x %>% filter(is_significant),
           aes(x = step_modded, color = Head),
           inherit.aes = FALSE,
           sides = "b",
           alpha = 0.8,
           length = unit(0.05, "npc")) +
  scale_x_log10() +
  geom_vline(xintercept = 1000, linetype = "dotted", size = 1.2) +
  geom_hline(yintercept = 0, linetype = "dashed", size = 1.2) +
  scale_color_viridis_d(option = "mako") +
  labs(x = "Training Step (Log10)",
       y = "Attention Difference",
       color = "Head") +
  facet_wrap(~Head, ncol = 2) +  # separate panel per head
  theme_minimal() +
  theme(text = element_text(size = 15),
        legend.position = "none")



### Stacked rugs
df_rugs <- df_attention_over_time %>%
  filter(mpath == "EleutherAI/pythia-410m") %>%
  inner_join(t_test_results) %>%
  filter(Layer == 1, p_adjusted < 0.05) %>%
  mutate(Head = factor(Head),
         rug_ymin = -0.25 - as.numeric(Head) * 0.02,  # more spacing here
         rug_ymax = rug_ymin + 0.01)  


df_attention_over_time %>%
  filter(mpath == "EleutherAI/pythia-410m") %>%
  inner_join(t_test_results) %>%
  filter(Layer == 1) %>%
  mutate(is_significant = p_adjusted < .05,
         Head = factor(Head)) %>%
  ggplot(aes(x = step_modded,
             y = mean_attention_diff,
             color = Head)) +
  geom_line(size = 1.2, alpha = 0.8) +
  geom_point(alpha = 0.5) +

  geom_segment(data = df_rugs,
               aes(x = step_modded, xend = step_modded,
                   y = rug_ymin, yend = rug_ymax, color = Head),
               inherit.aes = FALSE,
               alpha = 0.8) +

  scale_x_log10() +
  geom_vline(xintercept = 1000, linetype = "dotted", size = 1.2) +
  geom_hline(yintercept = 0, linetype = "dashed", size = 1.2) +
  scale_color_viridis_d(option = "mako") +
  labs(x = "Training Step (Log10)",
       y = "Attention Difference",
       color = "Head") +
  theme_minimal() +
  theme(text = element_text(size = 15),
        legend.position = "bottom")




### Layer 6
df_rugs <- df_attention_over_time %>%
  filter(mpath == "EleutherAI/pythia-410m") %>%
  inner_join(t_test_results) %>%
  filter(Layer == 6, p_adjusted < 0.05) %>%
  mutate(Head = factor(Head),
         rug_ymin = -0.25 - as.numeric(Head) * 0.02,  # more spacing here
         rug_ymax = rug_ymin + 0.01)  


df_attention_over_time %>%
  filter(mpath == "EleutherAI/pythia-410m") %>%
  inner_join(t_test_results) %>%
  filter(Layer == 6) %>%
  mutate(is_significant = p_adjusted < .05,
         Head = factor(Head)) %>%
  ggplot(aes(x = step_modded,
             y = mean_attention_diff,
             color = Head)) +
  geom_line(size = 1.2, alpha = 0.8) +
  geom_point(alpha = 0.5) +

  geom_segment(data = df_rugs,
               aes(x = step_modded, xend = step_modded,
                   y = rug_ymin, yend = rug_ymax, color = Head),
               inherit.aes = FALSE,
               alpha = 0.8) +

  scale_x_log10() +
  geom_vline(xintercept = 1000, linetype = "dotted", size = 1.2) +
  geom_hline(yintercept = 0, linetype = "dashed", size = 1.2) +
  scale_color_viridis_d(option = "mako") +
  labs(x = "Training Step (Log10)",
       y = "Attention Difference",
       color = "Head") +
  theme_minimal() +
  theme(text = element_text(size = 15),
        legend.position = "bottom")


```




# Attention vs. R2


## Load R2 data

```{r}
# setwd("/Users/seantrott/Dropbox/UCSD/Research/Ambiguity/SSD/entangled_meanings/src/analysis/")
directory_path <- "../../data/processed/rawc/pythia/distances/"
csv_files <- list.files(path = directory_path, pattern = "*.csv", full.names = TRUE)
csv_list <- csv_files %>%
  map(~ read_csv(.))
df_pythia_models_distance <- bind_rows(csv_list)

table(df_pythia_models_distance$mpath)

df_best_r2 = df_pythia_models_distance %>% 
  filter(mpath %in% c("EleutherAI/pythia-410m")) %>%
  group_by(mpath, revision, step, Layer, n_params) %>%
  summarise(r = cor(Distance, mean_relatedness, method = "pearson"),
            r2 = r ** 2) %>%
  group_by(mpath, step, n_params) %>%
  mutate(step_modded = step + 1) 
```

## Merge

```{r}
df_attention_over_time = df_pythia_models %>% 
  filter(mpath %in% c("EleutherAI/pythia-410m")) %>%
  group_by(mpath, revision, Layer, Head, step, n_params) %>%
  summarise(mean_attention = mean(Attention)) %>%
  mutate(step_modded = step + 1) %>%
  group_by(mpath, step, Head) %>%
  mutate(
    attention_diff = mean_attention - lag(mean_attention),
    Head = Head + 1
  )

df_merged_r2_attention = df_attention_over_time %>%
  inner_join(df_best_r2)


scale_factor <- max(df_merged_r2_attention$mean_attention, na.rm = TRUE) / 
                max(df_merged_r2_attention$r2, na.rm = TRUE)

df_merged_r2_attention %>%
  filter(Head == 2) %>%
  filter(Layer == 3) %>%
  ggplot(aes(x = step_modded)) +
  # Primary Y-axis: Attention
  geom_line(aes(y = mean_attention), size = 2, alpha = .7) +
  # Secondary Y-axis: R-squared
  geom_line(aes(y = r2 * scale_factor), 
            size = 2, linetype = "dotted") +
  scale_y_continuous(
    name = "Attention to Disambiguating Word",
    sec.axis = sec_axis(~ . / scale_factor, name = "R-squared")
  ) +
  scale_x_log10() +
  theme_minimal() +
  labs(
    x = "Training Step (Log10)",
    color = ""
  ) +
  # facet_wrap(~Head) +
  scale_color_viridis(option = "mako", discrete=TRUE) +
  theme(
    text = element_text(size = 15),
    legend.position = "bottom"
  )

```


## Analysis of attention and r2

```{r}
### Make actual best R2
df_best_r2_actual = df_best_r2 %>%
  group_by(revision) %>%
  slice_max(r2) %>%
  select(revision, step, r2)

### 
df_merged_r2_attention_actual = df_attention_over_time %>%
  inner_join(df_best_r2_actual)

# Pivot the dataframe to wider format
df_wide <- df_merged_r2_attention_actual %>%
  select(revision, Layer, Head, step, mean_attention, r2) %>%
  group_by(revision) %>%
  slice_max(r2) %>%
  pivot_wider(names_from = c(Layer, Head), 
              values_from = c(mean_attention), names_sep = "_") %>%
  ungroup()

### Target variable
y <- df_wide$r2

# Run separate regressions for each predictor and measure R^2 and coefficient
regression_results <- lapply(names(df_wide)[!(names(df_wide) %in% c("mpath", "revision", "step", "n_params", "step_modded", "r", "r2"))], function(var) {
  model <- lm(y ~ df_wide[[var]])
  summary_model <- summary(model)
  p_value <- coef(summary_model)[2, "Pr(>|t|)"]
  list(Feature = var, 
       R2 = summary_model$r.squared, 
       Coefficient = coef(model)[2],
       P_Value = p_value)
})


regression_results_df <- do.call(rbind, regression_results) %>% as.data.frame()

# Coerce types and apply multiple comparisons correction
regression_results_df <- regression_results_df %>%
  mutate(R2 = as.numeric(R2),
         Coefficient = as.numeric(Coefficient),
         P_Value = as.numeric(P_Value),
         P_Value_adj = p.adjust(P_Value, method = "fdr")) %>%
  separate(Feature, into = c("Layer", "Head"), sep = "_", convert = TRUE) 


regression_results_df %>%
  filter(Coefficient > 0) %>%
  filter(P_Value_adj < .05)

regression_results_df %>%
  mutate(Coefficient_sig = ifelse(P_Value_adj < 0.05, Coefficient, 0)) %>%
  filter(P_Value_adj < .05) %>%
  ggplot(aes(x = Layer, y = Head, fill = Coefficient_sig)) +
  geom_tile() +
  scale_fill_gradient2(low = "blue",
                       mid = "white",
                       high = "red",
                       midpoint = 0, 
                       name = "R2 ~ Attention") +
  theme_minimal() +
  labs(
    title = "",
    x = "Layer",
    y = "Head"
  )


t_test_results$is_sig = t_test_results$p_adjusted < .05

regression_results_df %>%
  mutate(Coefficient_floor = ifelse(Coefficient > 0, Coefficient, 0)) %>%
  inner_join(t_test_results) %>%
  filter(step == 143000) %>%
  mutate(
    t_test_floor = ifelse(t_statistic > 0, t_statistic, 0),
    label = ifelse(Coefficient_floor > 0 & t_test_floor > 0 & is_sig == TRUE,
                   paste(Layer, Head, sep = "_"),
                   NA)
  ) %>%
  ggplot(aes(x = Coefficient_floor,
             y = t_test_floor,
             color = is_sig)) +
  geom_point(size = 3, alpha = .5) +
  theme_minimal() +
  geom_hline(yintercept = 0 ,linetype = "dotted") +
  geom_text_repel(aes(label = label), size = 4) +
  geom_vline(xintercept = 0 ,linetype = "dotted") +
    scale_color_manual(values = viridisLite::viridis(2, option = "mako", 
                                                   begin = 0.8, end = 0.15)) +
  labs(x = "Coefficient (R2 ~ Attention)",
       y = "t-statistic (1-back analysis)") +
  theme(text = element_text(size = 15),
        legend.position = "none")
```
