---
title: "Pythia: Acquisition of Mod-NP Construction (v1)"
author: "Sean Trott and Pam Rivière"
date: "November 20, 2024"
output:
  html_document:
    keep_md: yes
    toc: yes
    toc_float: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(dpi = 300, fig.format = "pdf")
```


```{r include=FALSE}
library(tidyverse)
library(lmtest)
library(forcats)
library(broom)
library(lme4)
library(viridis)
library(ggridges)
library(lmerTest)
library(changepoint)

all_colors <- viridis::viridis(10, option = "mako")
my_colors <- all_colors[c(3, 5, 7)]  # Selecting specific colors from the palette
```


In this notebook, we analyze the trajectory of learning the "modifier NP" construction in Pythia models, which we analyze as the log ratio of an original modifier NP construction ("desperate act") vs. the reversed version ("act desperate"), each presented after an *end of sequence token* to each Pythia model; see V2 for an alternative version that presents the original and reversed versions in entire sentences (e.g., "It was a desperate act") and analyzes the relative log-probability of the modifier-NP phrases respectively.

# Load Pythia data


```{r}
# setwd("/Users/seantrott/Dropbox/UCSD/Research/Ambiguity/SSD/entangled_meanings/src/analysis/")
directory_path <- "../../data/processed/rawc/pythia/mod_np_cxn/"
csv_files <- list.files(path = directory_path, pattern = "*.csv", full.names = TRUE)
csv_list <- csv_files %>%
  map(~ read_csv(.))
df_pythia_models <- bind_rows(csv_list)

table(df_pythia_models$mpath)

```


# Analyzing log-ratio

## At final step

```{r}
df_pythia_models %>%
  # filter(mpath == "EleutherAI/pythia-14m") %>%
  filter(step %in% c(0, 1, 512, 1000, 2000, 143000)) %>%
  ggplot(aes(x = ratio,
             y = factor(step),
             fill = factor(mpath))) +
  geom_density_ridges2(aes(height = ..density..), 
                       color=gray(0.25), 
                       alpha = .5, 
                       scale=.85, 
                       # size=1, 
                       size = 0,
                       stat="density") +
  geom_vline(xintercept = 0, linetype = "dotted", size = 1.2) +
  labs(x = "Log-ratio (original vs. reversed)",
       y = "Step",
       fill = "") +
  theme_minimal() +
  scale_fill_viridis(option = "mako", discrete=TRUE) +
  theme(text = element_text(size = 15),
        legend.position="none")
```


## Across steps

```{r}
df_ratio_steps = df_pythia_models %>% 
  group_by(mpath, step, revision, n_params) %>%
  summarise(mean_lr_ratio = mean(ratio)) %>%
  mutate(step_modded = step + 1) %>%
  group_by(mpath) %>%
  mutate(
    mean_lr_ratio_diff = mean_lr_ratio - lag(mean_lr_ratio),
    mean_lr_ratio_diff2 = mean_lr_ratio_diff - lag(mean_lr_ratio_diff)
  ) %>%
  group_by(mpath) 

df_ratio_steps %>%
  ggplot(aes(x = step_modded,
             y = mean_lr_ratio,
             color = reorder(mpath, n_params))) +
  geom_point(size = 4, alpha = .7) +
  geom_line(size = 2, alpha = .7) +
  theme_minimal() +
  labs(x = "Training Step (Log10)",
       y = "Log ratio (original vs. reversed)",
       color = "") +
  geom_vline(xintercept = 1000, 
             linetype = "dotted", 
             size = 1.2) +
  scale_x_log10() +
  theme(text = element_text(size = 15),
        legend.position = "bottom") +
  scale_color_viridis(option = "mako", discrete=TRUE)


df_ratio_steps %>%
  ggplot(aes(x = step_modded,
             y = mean_lr_ratio_diff,
             color = reorder(mpath, n_params))) +
  geom_point(size = 4, alpha = .7) +
  geom_line(size = 2, alpha = .7) +
  theme_minimal() +
  geom_hline(yintercept = 0, linetype = "dotted", size = 2) +
  labs(x = "Training Step (Log10)",
       y = "Log ratio change (1st derivative)",
       color = "") +
  scale_x_log10() +
  geom_vline(xintercept = 1000, 
             linetype = "dotted", 
             size = 1.2) +
  theme(text = element_text(size = 15),
        legend.position = "bottom") +
  scale_color_viridis(option = "mako", discrete=TRUE)


df_ratio_steps %>%
  ggplot(aes(x = step_modded,
             y = mean_lr_ratio_diff2,
             color = reorder(mpath, n_params))) +
  geom_point(size = 4, alpha = .7) +
  geom_line(size = 2, alpha = .7) +
  theme_minimal() +
  geom_hline(yintercept = 0, linetype = "dotted", size = 2) +
  labs(x = "Training Step (Log10)",
       y = "Log ratio change (2nd derivative)",
       color = "") +
  scale_x_log10() +
  geom_vline(xintercept = 1000, 
             linetype = "dotted", 
             size = 1.2) +
  theme(text = element_text(size = 15),
        legend.position = "bottom") +
  scale_color_viridis(option = "mako", discrete=TRUE)




```


## Changepoint analysis for 14m

```{r}
df_14m = df_ratio_steps %>%
  filter(mpath == "EleutherAI/pythia-14m")

### get ratios per step
log_ratios <- df_14m$mean_lr_ratio
# Detect changepoints in the mean
cpt_result <- cpt.mean(log_ratios, method = "PELT")
# Extract components
changepoints <- cpts(cpt_result)             # Changepoint indices
means <- param.est(cpt_result)$mean          # Segment means
num_cpts <- ncpts(cpt_result)                # Number of changepoints

## Which training steps?
changepoint_steps <- df_14m$step[changepoints]
print(changepoint_steps)

## Annotate dataframe
df_14m <- df_14m %>%
  mutate(changepoint = ifelse(row_number() %in% changepoints, 1, 0))

df_14m %>%
  ggplot(aes(x = step_modded,
             y = mean_lr_ratio)) +
  geom_point(size = 4, alpha = .7) +
  geom_line(size = 2, alpha = .7) +
  theme_minimal() +
  labs(x = "Training Step (Log10)",
       y = "Log ratio (original vs. reversed)",
       color = "") +
  geom_vline(xintercept = changepoint_steps, color = "red", 
             linetype = "dashed", size = 1.2) +
  scale_x_log10() +
  theme(text = element_text(size = 15),
        legend.position = "bottom") +
  scale_color_viridis(option = "mako", discrete=TRUE)


```


